---
title: "MPSproto tutorial (advanced)"
#date: "01 08 2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## About MPSproto mixture interpretation
MPSproto is a R-package for interpreting MPS-STR mixtures. It requires a calibrated model in order to do so.
The user can choose between two models for the reads: Gamma or NegativeBinomial.
The differences are demonstrated in this tutorial.

### Step 1: Preparing data
```{r,fig.show='hide'}
library(MPSproto) #load package
pkg = path.package("MPSproto") #get package install folder
path = paste0(pkg,"/examples") #path of data to use

#Obtain calibrated model
calibratedModelFile = paste0(path,"/tutorialDataCalibrated.RDS") #file name for the calibrated model
calibration = readRDS(calibratedModelFile) #obtain the calibrated object

#We import the allele frequency
freqFile = paste0(path,"/ForenSeq_NIST1036cauc_FWbrack.csv") #frequency file to use
popFreq =  importMPSfreqs(freqFile)[[1]]

#import evidence and reference profiles
evidData = importMPSsample(paste0(path,"/MPS_evid2p.csv") ) #LOAD EVIDENCE DATA
refData = importMPSsample(paste0(path,"/MPS_evid2p_refs.csv") ) #LOAD REFERENCE DATA
plotMPS(evidData,refData) #Show in graphical interface

#Note: Ensure that the alleles are all in forward direction for all markers (not UAS)
```

### Step 2: Specify hypotheses for interpretation
Hypothesis set: minor as POI \
Hp: ref1 + 1 unknown \
Hd: 2 unknowns

```{r}
NOC = 2 #Number of contributors
fst0 = 0.01 #coancestry coefficient
hypHp = list(NOC=NOC,cond=c(1,0),fst=fst0) #(also include fst here)
hypHd = list(NOC=NOC,cond=c(0,0),fst=fst0,knownRef=1) #(also include fst here)

#If the major was evaluated as POI instead then
#hypHp = list(NOC=NOC,cond=c(0,1),fst=fst0) #(also include fst here)

```

### Step 3: Model fit based on gamma
```{r,fig.show='hide'}
mleHp = inferEvidence(evidData,popFreq,refData,hypHp , calibration, verbose=FALSE,model="GA")
mleHd = inferEvidence(evidData,popFreq,refData,hypHd , calibration, verbose=FALSE,model="GA")

#Obtain calculated LR
par = getPar(mleHp,mleHd) #get estimated parameters
LR = getLR(mleHp,mleHd) #get calculated LR (unlogged)
log10LR = getlog10LR(mleHp,mleHd) #log10 scale
log10LR
par

#Perform model validation
validhp = validMLE(mleHp,"Hp",verbose=FALSE)
validhd = validMLE(mleHd,"Hd",verbose=FALSE)

#Provide deconvolution
DCtableHp = deconvolve(mleHp)$table2
DCtableHd = deconvolve(mleHd)$table2

#Show model fit (expectation vs observations)
plotTopMPS(mleHp)
plotTopMPS(mleHd)
```




### Step 4: Model fit based on negative binomial
```{r,fig.show='hide'}
mleHp = inferEvidence(evidData,popFreq,refData,hypHp , calibration, verbose=FALSE,model="NB")
mleHd = inferEvidence(evidData,popFreq,refData,hypHd , calibration, verbose=FALSE,model="NB")

#Obtain calculated LR
par = getPar(mleHp,mleHd) #get estimated parameters
LR = getLR(mleHp,mleHd) #get calculated LR (unlogged)
log10LR = getlog10LR(mleHp,mleHd) #log10 scale
log10LR
par

#Perform model validation
validhp = validMLE(mleHp,"Hp",verbose=FALSE)
validhd = validMLE(mleHd,"Hd",verbose=FALSE)

#Provide deconvolution
DCtableHp = deconvolve(mleHp)$table2
DCtableHd = deconvolve(mleHd)$table2

#Show model fit (expectation vs observations)
plotTopMPS(mleHp)
plotTopMPS(mleHd)
```


