---
title: "MPSproto tutorial"
#date: "01 08 2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


## About MPSproto
MPSproto is a R-package for interpreting MPS-STR mixtures. It requires a calibrated model in order to do so.

The calibrated model ingredients must be: \
1) Analytical threshold parameters \
2) Marker efficiency parameters \
3) Noise parameters \
4) Stutter parameters (optional)

These parameters must be defined per marker and stored in a list structure (calibrated objected): \
*calibrated[[marker]]=list(markerEff,nNoiseParam,noiseSizeParam,AT,...)* \
where *...* is stutter type names (optional).


## Calibration
The data used for calibration in this tutorial are based on positive control only (reference 2800M).
Single source profiles must be provided where the genotype of the contributor is indicated.

### Step 1: Preparing data for calibration
```{r}
library(MPSproto) #load package
pkg = path.package("MPSproto") #get package install folder
path = paste0(pkg,"/examples") #path of data to use

#Load data for calibration;
calibrationDataFile = paste0(path,"/tutorialDataCalibration.RDS") #already stored in package
dat = readRDS(calibrationDataFile)  #load object

#note the header names must be strictly specified as follows:
colnames(dat) 
#Alternatively table data (text format) can be loaded using tableReader()

# Insert missing alleles with dose>0 (true alleles)
dat = insertMissingTrueAlleles(dat) 

#Specify Analytical threshold used (when exporting data)
AT=11  

#AT must be defined per marker:
locNames = unique(dat[,3])  #obtain locus names (Must be upper case)
if(length(AT)==1) AT = setNames( rep(AT,length(locNames)),locNames)

```

### Step 2: Calibrating marker efficiency
```{r}
markerEfficiency = inferMarkerEfficiency(dat,AT,verbose = FALSE) #fitting model MLE model
Amhat = markerEfficiency$MLE$theta2[locNames] #estimated params

```

### Step 3: Preparing stutter data (pre-step)
```{r}
#dat2 = getFilteredData(dat) #deprecated since already built into the getStutterData function
stutterData = getStutterData(dat, minStuttOccurence=5, verbose=FALSE) #obtaining stutter candidates (traversing different stutter types)
stuttDat = stutterData$stutterdata #obtain stutter data
```

### Step 4: Infer stutter model and store coefficients
```{r}
#A beta regression model is used to fit the stutter proportions
stuttMod = inferStutterModel(stuttDat)#,print2pdf="StutterModel")
stuttModTab = stuttMod$betaTable #obtain coefficient table

#Storing model coefficients in a list
stuttModList = list() #obtain list of coefficients
for(i in seq_len(nrow(stuttModTab))) {
  row = unlist(stuttModTab[i,])
  loc = row[1] #obtain locus name
  type = row[2] #obtain stutter type
  est =  as.numeric( na.omit(row[3:5]))    #obtain estimats
  if( !loc%in%names(stuttModList)) stuttModList[[loc]] = list()
  stuttModList[[loc]][[type]] = est
}

```


### Step 5: Calibrate noise model (done after stutter calibration)
```{r}
noiseMod = inferNoiseModel(sampleNames=unique(dat$SampleName), locNames=locNames, noisedata=stutterData$noisedata, AT=AT)#,print2pdf="NoiseModel")
```

### Step 6: Storing calibration object
```{r}
calibration = list()
for(loc in locNames) {
  markerEff= Amhat[loc] #setNames(c(,AmSD[loc],AmSD[loc]),c("est","SD","SD2"))
  noiseMods = lapply(noiseMod,function(x) x[loc])
  stuttMods = stuttModList[[loc]] 
  
  calibration[[loc]] = list(markerEff=markerEff)  
  calibration[[loc]] = append(calibration[[loc]],noiseMods)
  
  #put stutter model last (requirement!!)
  calibration[[loc]] = append(calibration[[loc]],stuttMods)
}

#Store object for later use (mixture interpretation)
saveRDS(calibration,"tutorialDataCalibrated.RDS")
```

### A model calibration wrapper function
```{r, eval=FALSE}
#It is possible to run the following wrapper function to perform model calibration (still under development):
calibrateModel(dat,AT=11,minStuttOccurence=5,verbose=FALSE)
```


## Mixture interpretation

### Step 1: Data import (calibrated model and frequencies)
```{r}
#The calibrated model is imported
calibratedModelFile = paste0(path,"/tutorialDataCalibrated.RDS") #file name for the calibrated model
calibration = readRDS(calibratedModelFile) #obtain the calibrated object

#We import the allele frequency
freqFile = paste0(path,"/ForenSeq_NIST1036cauc_FWbrack.csv") #frequency file to use
popFreq =  importMPSfreqs(freqFile)[[1]]

#Note: Ensure that the alleles are all in forward direction for all markers (not UAS)
```

### Step 2: Generating a fictive mixture dataset
```{r}
#SETTINGS FOR GENERATED SAMPLE
NOC = 2 #Number of contributors
fst0 = 0.01

set.seed(1) #make reproducible
gen = genMPSevidence(calibration,NOC,popFreq,mu=1000,omega=0.1)
evidData = gen$samples #this is evidence data
refData = gen$refData #this is reference data

#We can Store Sample to text file:
#write.csv2(sample_listToTable(evidData),file="MPS_evid2p.csv",row.names = FALSE)
#write.csv2(sample_listToTable(refData),file="MPS_evid2p_refs.csv",row.names = FALSE)

```


### Step 3: Import and visualize profiles
```{r}
#Samples can also be obtained from a text file:
evidData = importMPSsample(paste0(path,"/MPS_evid2p.csv") ) #LOAD EVIDENCE DATA
refData = importMPSsample(paste0(path,"/MPS_evid2p_refs.csv") ) #LOAD REFERENCE DATA
plotMPS(evidData,refData) #Show in graphical interface
```

### Step 4: Specify hypotheses for interpretation
Hypothesis set: minor as POI \
Hp: ref1 + 1 unknown \
Hd: 2 unknowns

```{r}
hypHp = list(NOC=NOC,cond=c(1,0),fst=fst0) #(also include fst here)
hypHd = list(NOC=NOC,cond=c(0,0),fst=fst0,knownRef=1) #(also include fst here)

#If the major was evaluated as POI instead then
#hypHp = list(NOC=NOC,cond=c(0,1),fst=fst0) #(also include fst here)
#hypHd = list(NOC=NOC,cond=c(0,0),fst=fst0,knownRef=2) #(also include fst here)

```

### Step 5: Model fit of Hp and Hd 
```{r}
mleHp = inferEvidence(evidData,popFreq,refData,hypHp , calibration, verbose=FALSE)
mleHd = inferEvidence(evidData,popFreq,refData,hypHd , calibration, verbose=FALSE)
```

### Step 6: Obtain calculated LR
```{r}
par = getPar(mleHp,mleHd) #get estimated parameters
LR = getLR(mleHp,mleHd) #get calculated LR (unlogged)
log10LR = getlog10LR(mleHp,mleHd) #log10 scale
upperLR = getUpperLR(mleHd) #get upper LR-limit 1/RMP(theta)
```

### Step 7: Perform model validation
```{r,fig.show='hide',eval=FALSE}
validhp = validMLE(mleHp,"Hp")
validhd = validMLE(mleHd,"Hd")
```

### Step 8: Provide deconvolution
```{r}
DCtableHp = deconvolve(mleHp)$table2
DCtableHd = deconvolve(mleHd)$table2
```

### Step 9: Show model fit (expectation vs observations)
```{r}
plotTopMPS(mleHp)
plotTopMPS(mleHd)
```


